Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2024-01-29 manuscript_revised.RMD
updating revised manuscript

# Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

@@

::: {custom-style="correspondence_style"}
Correspondance: Kristian Lian ([kristian.lian\@inn.no](mailto:kristian.lian@inn.no){.email}), Stian Ellefsen ([stian.ellefsen\@inn.no](mailto:stian.ellefsen@inn.no){.email})
:::

@@

<!-- To do: Fix tables, add outlier figures to the readme -->

@@

**Purpose:** Resistance training-induced skeletal muscle hypertrophy seems to depend on ribosome biogenesis and content. High glucose treatment may augment ribosomal biogenesis, as evident from *in vitro* studies, and may thus act to potentiate resistance training-induced adaptations *in vivo*. This was investigated with total RNA and ribosomal RNA abundances as main outcomes, with relevant transcriptional or translational regulators (c-Myc/UBF/rpS6) as a secondary outcome.

**Methods:** Sixteen healthy, moderately trained individuals (male/female, n = 9/7; age, 24.1 (3.3)), participated in a within-participant crossover trial, conducting unilateral resistance training (leg press and knee extension, 3 sets of 10 repetitions maximum) with pre- and post-exercise ingestion of either glucose (3x30g, 90g total) or placebo supplement (Stevia rebaudiana extract, 3x0.3g, 0.9g total), together with protein (2x25g, 50g total), on alternating days for twelve days; six sessions per condition, resistance training was performed in an otherwise fasted state, as morning sessions. Micro biopsies were sampled from m. vastus lateralis before and after the intervention, and maximal unilateral isometric and isokinetic knee extension torque were measured before, during (days 4/5, 8/9) and after the intervention.

**Results:** Glucose ingestion did not have beneficial effects on resistance training-induced increases of ribosomal content (mean difference 7.6% [-7.2, 24.9], *p* = 0.34; ribosomal RNA, 47S/18S/28S/5.8S/5S, range 7.6%-37.9%, *p* = 0.40-0.98) or levels of relevant transcriptional or translational regulators (c-MYK/UBF/rpS6, *p* = 0.094-0.292). Of note, both baseline and trained state data of total RNA showed a linear relationship with UBF; a ∼14% increase in total RNA corresponded to 1 standard deviation unit increase in UBF (*p* = 0.003).

**Conclusion:** Glucose ingestion before and after resistance training did not augment ribosomal RNA accumulation during twelve days of heavy-load resistance training in moderately trained young adults.

@@

Gratitude is due, to Henriette Spilhaug Bollandsås, Nora Lerdalen Bonsak and Thea Bøhn Nyløkken for their dedication and hard work during the training intervention. The authors would also like to express their gratitude to all the participants for staying committed to this demanding protocol in an impressive way. We also thank Håvard Nygaard for lending your expertise and technical skills to improve this project.

@@

Responses to systematic resistance training (RT) vary widely between individuals, with as much as 10-15% showing impaired skeletal muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. There may be differences in the genetic predisposition in terms of readiness to adapt to exercise [@thalacker-mercer_cluster_2013], however, the internal physiological milieu seems favourably shaped for muscle growth by factors such as different types of training stress and nutrients [@figueiredo_genetic_2021; @sebata_control_2018; @thalacker-mercer_cluster_2013]. Indeed, training stress and nutrients impact the ability to synthesise ribosomes, which in turn is connected to the magnitude of RT-induced responses in an individual [@hammarstrom_benefits_2020; @kusnadi_regulation_2015; @sebata_control_2018]. For instance, increasing training volume generally induces greater ribosome biogenesis and thus, greater benefits of RT [@hammarstrom_benefits_2020; @krieger_single_2009; @schoenfeld_strength_2017]. Still, as evident from Hammarström et al. (2020), not all participants experience increased muscle mass and -strength despite increasing training volume. Therefore, other means than modification of RT variables alone seem necessary to optimise individual responses to RT.

Nutritional supplements such as protein and creatine are often used to optimise RT adaptations [@cermak_protein_2012; @lanhers_creatine_2015; @lanhers_creatine_2017; @morton_systematic_2018]. However, it remains equivocal if other nutritional adjuvants such as glucose can increase the efficacy of RT [@tezze_anabolic_2023]. This is surprising since glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major energy supplier to cells via adenosine triphosphate (ATP) synthesis [@mul_exercise_2015; @sebata_control_2018]. Furthermore, energy availability is a decisive factor in the *de novo* synthesis of ribosomes [@kusnadi_regulation_2015; @moss_housekeeper_2007; @sebata_control_2018] which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019; @hammarstrom_benefits_2020; @sebata_control_2018; @stec_ribosome_2016; @von_walden_ribosome_2019]. In addition, insulin may have anabolic effects with elevated amino acids, and has a clear role in the reduction of muscle protein breakdown independent of amino acid availability [@abdulla_role_2016; @hillier_physiological_2000]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the potential additive effect of glucose and RT on ribosome biogenesis.

Ribosomal biogenesis and content seem to be a prerequisite for skeletal muscle growth, and transcription of ribosomal ribonucleic acid (rRNA) by RNA Polymerase I is considered the rate-limiting step in synthesising new ribosomes [@moss_promotion_1995]. Multiple signalling pathways converge to regulate rRNA transcription, including c-Myc and the mammalian target of rapamycin complex 1 (mTORC1) signal-transduction pathway [@kusnadi_regulation_2015; @von_walden_ribosome_2019; @sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. The general transcription factor c-Myc increases ribosomal biogenesis directly through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011; @mori_c-myc_2021; @west_acute_2016]. Indeed, UBF phosphorylation, which is required for interaction with the ribosomal deoxyribonucleic acid (rDNA) promoter, seems to be increased by high-glucose treatment in a mTORC1-dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells [@mariappan_ribosomal_2011]. In addition, mTORC1 regulation of ribosomal gene transcription is dependent on ribosomal protein S6 kinase beta-1 (S6K1) [@hannan2003mtor] with its target substrate ribosomal protein S6 (rpS6) also increasing during early phase RT in humans [@hammarstrom_ribosome_2022]. Independent of UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promoted rRNA transcription in cell cultures [@zhai_human_2012]. The mTORC1 pathway receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009] and contributes to ribosome biogenesis through the forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription and through the regulation of ribosomal protein translation [@figueiredo_regulation_2019; @von_walden_ribosome_2019]. Moreover, mTORC1 and S6K1 serves as direct mediators of insulin signalling in skeletal muscle [@hillier_physiological_2000]. Together these observations indicate a potential role of glucose in positively affecting human skeletal muscle rRNA synthesis, through potentiating transcription or translation of ribosomal RNA and increasing the translational capacity of skeletal muscle cells [@sebata_control_2018; @zhai_human_2012; @hillier_physiological_2000; @hoppe_amp-activated_2009].

There have been multiple studies recently suggesting translational capacity to be as important, if not more important than, translational efficiency concerning long-term skeletal muscle adaptations to RT [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022; @figueiredo_revisiting_2019]. Previously, *in vitro* studies have shown UBF and c-Myc to be important factors in the PIC, c-Myc as a general transcription factor and UBF as a specific transcription factor for rDNA transcription initiation [@mariappan_ribosomal_2011; @von_walden_ribosome_2019]. Coupled with the recent observations by Hammarström et al. [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022], understanding underlying factors regulating and affecting ribosome biogenesis seems key to furthering our understanding of the optimisation of RT to individual phenotypes.

Therefore, the purpose of this investigation was to test the hypothesis that glucose supplementation before and after five RT sessions over 12 days will potentiate RT-associated accumulation of markers of ribosomal abundance following five RT sessions. Secondly, we aimed to describe the association between changes in total RNA abundance and UBF in human skeletal muscle.

@@

All participants gave their written informed consent before study enrollment. The study was approved by the regional ethical committee (REK, ID nr. 153628), pre-registered at clinicaltrials.gov (Identifier: NCT04545190), and conducted according to the Helsinki Declaration.

@@

Sixteen healthy male and female participants (20-33 years, Table 1) were recruited to the study through social media advertisement and word of mouth. The eligibility criteria were non-smokers and moderately trained (i.e. 2-8 RT sessions per 14 days for the last six months). Exclusion criteria were previous injury leading to impaired strength, inability to perform resistance exercise training, symptoms, and a medical record of metabolic disorders including hyperglycemia. Of the sixteen participants that commenced the intervention, three participants dropped out. One due to sickness and inability to resume, two participants experienced muscular discomfort related to heavy resistance training. Lean mass (Table 1) was measured using Dual-Energy X-ray Absorptiometry (DXA, Prodigy Advance PA+302047, Lunar, San Francisco, CA, USA) on Day -1, the last day preceding the RT intervention.

@@

The study was designed as a 12-day double-blinded placebo-controlled simultaneous crossover trial, with an alternating unilateral RT protocol (Figure 1A). Participants were randomly allocated to exercise one leg with a glucose condition and one leg with a placebo condition (Figure 1A). One person was exclusively responsible for the randomisation code and supplement distribution, blinding both investigators and participants regarding which leg exercised with glucose/placebo conditions. Glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (Steviosa, Soma Nordic AS, Oslo, Norway) were masked by mixing with 300ml Fun Light (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g glucoseum monohydricum vs. \~0.3g Stevia rebaudiana extract). In this blinded taste test, the participants were given two boluses of placebo (75ml) and two boluses of glucose (75ml), consumed in a randomised order. The participants were instructed to finish one bolus, guess the content, and move on to the next bolus. To ensure equal conditions during training sessions and strength testing, participants exercised and tested at the same time of day, +/- 1hr with the same supervisor on pairwise consecutive days (i.e. on days 1-2, 3-4, etc.). To further standardise this, participants also recorded and repeated their daily macronutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

All participants completed six RT sessions with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose and serum c-peptide levels). Participants were asked to avoid resistance- or high-intensity training of the legs from Day -7 (Figure 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from external exercise sources.

@@

The dietary intervention spanned the whole day, divided into three periods: I) From awakening until 2.5 hours (hrs) after RT, II) from 2.5 hours after RT until 2200hrs, and III) from 2200hrs until awakening. During period I, participants ingested protein and glucose *or* protein and placebo only. Glucose/placebo was ingested three times in period I: 30 minutes before RT (0830hrs, 30g vs. 0g glucose), immediately before RT (0900hrs, 30g vs. 0g glucose), and immediately after RT (\~0930hrs, 30g vs. 0g glucose). Whey Protein Isolate (Proteinfabrikken, Stokke, Norway) was ingested 2hrs before RT (0700hrs) and immediately after RT (\~0930hrs), in boluses of 25 g mixed with 150 ml water. In the afternoon (1800hrs-1900hrs, period II) participants ingested glucose or placebo (3x30g vs. 3x0g glucose) opposite to the supplement they received during RT, to ensure a balanced daily intake of glucose. Apart from this, participants ingested a self-chosen diet during period II. Further, participants were asked not to use any other supplements such as additional protein and/or creatine, and to register all food/drink consumption in MyFitnessPal or similar applications. The self-chosen diet was repeated on pairwise consecutive days, to ensure similar premises for resistance training responses between conditions. During period III (2200-0700hrs), participants remained in an overnight fasted state. The daily onset of the dietary intervention (i.e., first ingestion of protein supplement) varied between participants, from 0600hrs to 0900hrs to allow multiple participants to complete the protocol simultaneously. During sessions, participants were free to ingest water ad libitum.

@@

Strength tests were performed before (Figure 1A, Days -7 and -5, and -1, both legs) and during the intervention (Figure 1A, on days 4 and 8 for leg 1, and days 5 and 9 for leg two), after session 5 and after finalization of the intervention (Figure 1A, on days 11/12 for leg 1 and days 12/13 for leg 2). Maximal isometric and isokinetic knee extension torque was measured with a Humac Norm Dynamometer (CSMi, Stoughton, Massachusetts, USA). Individual positions were recorded and standardized from pre-intervention tests (Figure 1A, days -7 and -5). Isokinetic peak torque was measured concentrically from 90 to 0 degrees knee angle (extended knee was set to 0 degrees) at angular velocities of 60- and 240 degrees per second, 2x3 repetitions each, with the first set of each exercise as a sub-maximal warm-up. Isometric knee extensor peak torque was measured at a knee angle of 60 degrees, for a maximum of 10 seconds and two repetitions per test. The isometric tests were ended when the participants reached a plateau or peak torque development decreased, which on average occurred between 2-4 seconds into the test. During days 4, 5, 8 and 9 (Figure 1D, days 4 and 5 = Post 2RT, days 8 and 9 = Post 4RT), humac tests were conducted one hour before RT on the leg performing RT the previous day. During days 11 and 12, humac tests were performed four times: I) 45min before the last RT session (Figure 1D, Post 5RT), II) 30min after the last RT session (Figure 1D, 30 min post 6RT), III) 2hrs after the last RT session (Figure 1D, 2h post 6RT), and IV) 23hrs after the last RT session (Figure 1D, 23h post 6RT). Test I on day 12/13 included testing of both legs, representing 23hrs post-RT session test of one leg and post-session 5 test of the other leg. The highest peak torque values from the respective angular velocities and time points were summarized in an index. The index was calculated by dividing the average peak torque value by the highest observed peak torque value per angular velocity and summarizing this new index per angular velocity to a mean strength index.

Assessment of unilateral one repetition maximum (1RM) leg press and knee extension was conducted at before the intervention during familiarization (Figure 1, Days -7 and -5). The participants performed a general warm-up with 10 minutes of cycling on an indoor exercise bicycle. A protocol consisting of 1x10, 1x6 and 1x3 repetitions with a load equivalent to \~50-75% of assumed max repetitions, was used as a specific warm-up before each of the tests. All positions were controlled and recorded at the first 1RM test and repeated for the RT sessions. Maximal leg press strength was defined as the maximal load lifted in a controlled fashion, starting from a knee angle of 90 degrees. To find a reproducible 90-degree knee angle for each participant, centimetre markings on the side panels of the leg press machine were used to record where to find 90 degrees for each separate leg and participant. Attempts where participants did not reach 90 degrees during the eccentric phase, were not approved. Maximal knee extension testing followed the same specific warm-up as the maximal leg press test and was defined as the maximal load lifted in a controlled fashion, reaching full extension of the knee joint. Attempts with exaggerated hip movement or beneath full extension were not approved. Two minutes of rest were given during the specific warm-up, and three minutes of rest were given between 1RM attempts.

@@

Resistance training consisted of three sets of unilateral leg presses and three sets of unilateral knee extensions, with an intensity of 10 repetitions maximum (10RM). As a general warm-up, the participants cycled on an indoor exercise bicycle for 5-10 minutes. In addition, before the respective exercises, two 10-repetition warm-up sets were completed at \~50% and \~70% of 10RM. To ensure adequate exercise stimulation throughout the intervention, the exercise load was increased the following set if the participants lifted more than 12 repetitions, as a progressive loading strategy. If the participants lifted fewer than 8 repetitions per set, the load was reduced in the following set. The resting time between working sets was two minutes. For safety and standardisation purposes, all sessions were monitored by trained personnel. Lastly, training volume (load and repetitions) was logged for every session.

@@

To measure blood glucose levels with and without glucose ingestion/training, capillary blood was collected from finger draws on days with biopsy sampling. One capillary blood sample was collected on day 1 (Figure 1A) to serve as a baseline. On days 11 and 12 (Figure 1A), capillary blood samples were collected seven times: I) Immediately before protein ingestion (0700hrs) II) 45 minutes after protein ingestion (0745hrs) III) 1.5 hours after protein ingestion (0830hrs, i.e., immediately before GLU/PLAC intake), IV) 2 hours after protein ingestion (0900hrs, i.e., immediately before training), IV) in the middle of RT (\~0915hrs), V) immediately after training (\~0930hrs), and VI) 2 hours after completion of training (\~1130hrs). Capillary blood samples were analysed with in-house equipment (BIOSEN C-Line, EKF diagnostic GmbH, Barleben). Venous blood samples were collected from the antecubital vein, coinciding with the capillary samples except 45 minutes after protein ingestion and in the middle of the RT session, to analyse endocrine variables.

>>>>

# Glucose ingestion before and after resistance training sessions does not augment ribosome biogenesis in healthy moderately trained young adults

@@

::: {custom-style="correspondence_style"}
Correspondance: Kristian Lian ([kristian.lian\@inn.no](mailto:kristian.lian@inn.no){.email})
:::

@@

@@

**Purpose:** Resistance training-induced skeletal muscle hypertrophy seems to depend on ribosome biogenesis and -content. High glucose treatment may augment ribosome biogenesis, as evident from *in vitro* studies, and may thus act to potentiate resistance training-induced adaptations *in vivo*. This was investigated with total RNA and ribosomal RNA as main outcomes, with relevant transcriptional or translational regulaters (c-Myc/UBF/rpS6) as a secondary outcome.

**Methods:** Sixteen healthy, moderately trained individuals (male/female, n = 9/7; age, 24.1 (3.3)) participated in a within-participant crossover trial where they conducted unilateral resistance training (leg press and knee extension, 3 sets of 10 repetitions maximum) with pre- and post-exercise ingestion of either glucose (3x30g, 90g total) or placebo supplements (Stevia rebaudiana, 3x0.3g, 0.9g total), together with protein (2x25g, 50g total), on alternating days for twelve days. Six morning resistance exercise sessions were conducted per condition, and the sessions were performed in an otherwise fasted state. Micro biopsies were sampled from m. vastus lateralis before and after the intervention, and maximal unilateral isometric and isokinetic knee extension torque were measured before, during (days 4/5, 8/9) and after the intervention.

**Results:** Glucose ingestion did not have beneficial effects on resistance training-induced increases of ribosomal content (mean difference 7.6% [-7.2, 24.9], *p* = 0.34; ribosomal RNA, 47S/18S/28S/5.8S/5S, range 7.6%-37.9%, *p* = 0.40-0.98) or levels of relevant transcriptional or translational regulators (c-MYK/UBF/rpS6, *p* = 0.094-0.292). Of note, both baseline and trained state data of total RNA showed a linear relationship with UBF; a ∼14% increase in total RNA corresponded to 1 SD unit increase in UBF (*p* = 0.003).

**Conclusion:** Glucose ingestion before and after resistance training sessions did not augment ribosomal RNA accumulation during twelve days of heavy-load resistance training in moderately trained young adults.

@@

Gratitude is due, to Henriette Spilhaug Bollandsås, Nora Lerdalen Bonsak, Thea Bøhn Nyløkken, Max Ullrich and Chris Sylstad for their dedication and hard work during the training intervention. The authors would also like to express their gratitude to all the participants for staying committed to this demanding protocol in an impressive way. We also thank Håvard Nygaard for lending your expertise and technical skills to improve this project.

@@

Responses to systematic resistance training (RT) vary widely between individuals, with as much as 10-15% showing impaired skeletal muscle growth in response to standardised training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. Genetic predisposition may explain some of this variation [@thalacker-mercer_cluster_2013], but in general the internal physiological milieu seems to be favourably shaped for muscle growth in response to different types of nutrient intake and exercise training [@figueiredo_genetic_2021; @sebata_control_2018; @thalacker-mercer_cluster_2013]. Indeed, exercise training and nutrients impact the ability to synthesise ribosomes, which in turn is demonstrated to be connected to the magnitude of RT-induced responses [@hammarstrom_benefits_2020; @kusnadi_regulation_2015; @sebata_control_2018]. For instance, increasing training volume generally induces greater ribosome biogenesis and is associated with greater benefits of RT in terms of gain in skeletal muscle mass and -strength [@hammarstrom_benefits_2020; @krieger_single_2009; @schoenfeld_strength_2017]. Still, as evident from Hammarström et al. (2020), not all participants experience increased muscle mass and improved muscle strength with increasing training volume. Therefore, means other than modification of RT variables seem necessary to optimise individual responses to RT, for instance nutritional adjuvants.

Nutritional supplements such as protein and creatine are frequently advocated as means to optimise RT adaptations [@cermak_protein_2012; @lanhers_creatine_2015; @lanhers_creatine_2017; @morton_systematic_2018]. The efficacy of other nutritional adjuvants such as glucose remains equivocal [@tezze_anabolic_2023]. This is surprising given that glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major energy supplier to cells via adenosine triphosphate (ATP) synthesis [@mul_exercise_2015; @sebata_control_2018]. Furthermore, energy availability is a decisive factor in the *de novo* synthesis of ribosomes [@kusnadi_regulation_2015; @moss_housekeeper_2007; @sebata_control_2018] which in turn seems to determine muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019; @hammarstrom_benefits_2020; @sebata_control_2018; @stec_ribosome_2016; @von_walden_ribosome_2019]. In addition, insulin, which is secreted from the beta cells of pancreas in response to rising blood glucose levels, may itself exert anabolic effects irrespective of muscle contraction, e.g. by elevating levels of amino acids, an also plays a role in reducing muscle protein breakdown independent of amino acid availability [@abdulla_role_2016; @hillier_physiological_2000]. It seems plausible that combined RT and glucose ingestion provides effects on ribosome biogenesis compared to RT alone.

Ribosome biogenesis and content seem to be a prerequisite for skeletal muscle growth, and transcription of ribosomal ribonucleic acid (rRNA) by RNA Polymerase I is considered the rate-limiting step in *de novo* ribosome biogenesis [@moss_promotion_1995]. Multiple proteins and signalling pathways converge to regulate rRNA transcription, including c-Myc and the mammalian target of rapamycin complex 1 (mTORC1) signal-transduction pathway [@kusnadi_regulation_2015; @von_walden_ribosome_2019; @sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. First, the general transcription factor c-Myc increases ribosome biogenesis directly through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011; @mori_c-myc_2021; @west_acute_2016]. Indeed, UBF phosphorylation, which is required for interaction with the ribosomal deoxyribonucleic acid (rDNA) promoter, is increased by RT alone in muscle and also seems to be increased by high-glucose treatment in a mTORC1-dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells [@mariappan_ribosomal_2011].

Second, the mTORC1 pathway receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009]. This contributes to ribosome biogenesis by forming the preinitiation complex (PIC) that marks the initiation of rRNA transcription, as well as through regulation of ribosomal protein translation [@figueiredo_regulation_2019; @von_walden_ribosome_2019]. In addition, mTORC1 and ribosomal protein S6 kinase beta-1 (S6K1) are direct mediators of insulin signalling in skeletal muscle [@hillier_physiological_2000]. Third, high glucose was shown to lead to chromatin remodelling independent of UBF and mTORC1, which in turn promotes rRNA transcription in cell cultures [@zhai_human_2012]. Together, these mechanistic observations underscore a potential role for glucose in muscle ribosome biogenesis and function in human skeletal muscle, acting in concert with RT to potentiate transcription of ribosomal RNA and increasing translational capacity [@sebata_control_2018; @zhai_human_2012; @hillier_physiological_2000; @hoppe_amp-activated_2009].

Multiple studies have suggested translational capacity to be as important, if not more important, than translational efficiency for promoting long-term skeletal muscle adaptations to RT [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022; @figueiredo_revisiting_2019]. While the regulation of translational capacity itself involves activation of c-Myc and UBF, acting to stimulate formation of the PIC through the general transcription factor, as well as through a specific transcription factor facilitating rDNA transcription initiation [@mariappan_ribosomal_2011; @von_walden_ribosome_2019], the content of ribosomes, c-Myc and UBF is increased with RT-induced muscle accretion [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022]. Furthermore, Nakada et al. [@nakada2016correlation] found a correlation between rRNA content and rpS6 content in their synergist ablation model on rats. This makes knowledge about factors that regulate and affect ribosome biogenesis key for optimising RT at the individual level.

Therefore, the main purpose of this investigation was to test the hypothesis that glucose supplementation given before and after six RT sessions, conducted over a period of 12 days will potentiate RT-associated accumulation of markers of ribosomal abundance. Secondly, we aimed to describe the association between changes in total RNA abundance and UBF in human skeletal muscle.

@@

All participants gave their written informed consent before study enrolment. The study was approved by the Regional Committee for Medical and Health Research Ethics - South-East Norway (ID nr. 153628), pre-registered at clinicaltrials.gov (Identifier: NCT04545190), and conducted according to the Helsinki Declaration.

@@

Sixteen healthy male and female participants (20-33 years, Table 1) were recruited to the study through social media advertisement and word of mouth. The eligibility criteria were non-smoking and moderately trained (i.e. 2-8 RT sessions per 14 days for the last six months). Exclusion criteria were previous injury leading to impaired muscle strength, inability to perform resistance exercise training, symptoms, and a medical record of metabolic disorders including hyperglycemia. Of the sixteen participants that commenced the intervention, three participants dropped out. One due to sickness and inability to resume, while two participants experienced muscular discomfort related to heavy resistance training. Lean mass and body fat % (Table 1) was measured using Dual-Energy X-ray Absorptiometry (DXA, Prodigy Advance PA+302047, Lunar, San Francisco, CA, USA) on Day -1, the last day preceding the RT intervention.

Table 1 could be placed here.

@@

The study was designed as a 12-day double-blinded placebo-controlled simultaneous crossover trial, with an alternating unilateral RT protocol (Figure 1A). Participants were randomly allocated to exercise one leg with a glucose condition and one leg with a placebo condition (Figure 1A). One person was exclusively responsible for the randomisation code and supplement distribution, blinding both investigators and participants regarding which leg exercised with glucose/placebo conditions. One bolus of glucose was ingested as 30g of glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and one bolus of placebo was ingested as 100mg Stevia powder (Steviosa, Soma Nordic AS, Oslo, Norway), containing the natural sweetener erythritol in amounts equivalent to the sweetness of 30g glucose, mixed with 300ml Fun Light (Orkla, Oslo, Norway). Hence, the glucose and placebo supplements had identical chemical composition, except for their content of glucose/Stevia. To test whether the boluses truly were masked sufficiently to avoid detection, a blinded taste test was carried out. In this blinded taste test, the participants were given two glasses of glucose mix (75ml per) and two glasses of the placebo mix (75ml per), consumed in a randomised order per participant. The participants were instructed to finish one bolus, note their guess for its content, and move on to the next glass. On average, the participants had a score of 2 points (2.08 ± 1.24) out of 4 possible. To ensure equal conditions during training sessions and strength testing, participants exercised and tested at the same time of day, +/- 1hr with the same supervisor on pairwise consecutive days (i.e. on days 1-2, 3-4, etc.). To further standardise this, participants also recorded and repeated their daily macronutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

All participants completed six RT sessions with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose and plasma c-peptide levels). Participants were asked to avoid resistance- or high-intensity training of the legs from Day -7 (Figure 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from external exercise sources.

@@

The dietary intervention spanned the whole day, divided into three periods: I) From awakening until 2.5 hours (hrs) after RT, II) from 2.5 hours after RT until 2200hrs, and III) from 2200hrs until awakening. During period I, participants ingested protein and glucose *or* protein and placebo only. Glucose/placebo was ingested three times in period I: 30 minutes before RT (30g vs. 0g glucose), immediately before RT (30g vs. 0g glucose), and immediately after RT (30g vs. 0g glucose). Whey Protein Isolate (Proteinfabrikken, Stokke, Norway) was ingested 2hrs before RT and immediately after RT, in boluses of 25 g mixed with 150 ml water. In the afternoon (1800-1900hrs, period II) participants ingested glucose or placebo (3x30g vs. 3x0g glucose) opposite to the supplement they received during RT, to ensure a balanced daily intake of glucose. Apart from this, participants ingested a self-chosen diet during period II. Further, participants were asked not to use any other supplements such as additional protein and/or creatine, and to register all food/drink consumption in MyFitnessPal or similar applications. The self-chosen diet was repeated on pairwise consecutive days, to ensure similar premises for resistance training responses between conditions. During period III (2200-0700hrs), participants remained in an overnight fasted state. The daily onset of the dietary intervention (i.e., first ingestion of protein supplement) varied between participants, from 0600hrs to 0900hrs to allow multiple participants to complete the protocol simultaneously. During resistance training sessions, participants were free to ingest water ad libitum.

@@

Muscle strength tests were performed before (Figure 1A, Days -7 and -5, and -1, both legs) and during the intervention (Figure 1A, on days 4 and 8 for leg 1, and days 5 and 9 for leg two), after session 5 and after finalization of the intervention (Figure 1A, on days 11/12 for leg 1 and days 12/13 for leg 2). Maximal isometric and isokinetic knee extension torque was measured with a Humac Norm Dynamometer (CSMi, Stoughton, Massachusetts, USA). Individual positions were recorded and standardized from pre-intervention tests (Figure 1A, days -7 and -5). Isokinetic peak torque was measured concentrically from 90 to 0 degrees knee angle (extended knee was set to 0 degrees) at angular velocities of 60- and 240 degrees per second, 2x3 repetitions each, with the first set of each exercise as a sub-maximal warm-up. Isometric knee extensor peak torque was measured at a knee angle of 60 degrees, for a maximum of 10 seconds and two repetitions per test. The isometric tests were ended when the participants reached a plateau or peak torque development decreased, which on average occurred between 2-4 seconds into the test. During days 4, 5, 8 and 9 (Figure 1D, days 4 and 5 = Post 2RT, days 8 and 9 = Post 4RT), knee extension torque tests were conducted one hour before RT on the leg performing RT the previous day. During days 11 and 12, these tests were performed four times: I) 45min before the last RT session (Figure 1D, Post 5RT), II) 30min after the last RT session (Figure 1D, 30 min post 6RT), III) 2hrs after the last RT session (Figure 1D, 2h post 6RT), and IV) 23hrs after the last RT session (Figure 1D, 23h post 6RT). Test I on day 12/13 included testing of both legs, representing 23hrs post-RT session test of one leg and post-session 5 test of the other leg. The highest peak torque values from the respective angular velocities and time points were summarized in an index. The index was calculated by dividing the average peak torque value by the highest observed peak torque value per angular velocity and summarizing this new index per angular velocity to a mean strength index.

Assessment of unilateral one repetition maximum (1RM) leg press and knee extension was conducted in the familiarisation phase prior to the intervention (Figure 1, Days -7 and -5). The participants performed a general warm-up with 10 minutes of cycling on an indoor exercise bicycle. A protocol consisting of 1x10, 1x6 and 1x3 repetitions with a load equivalent to \~50-75% of assumed maximal repetitions, was used as a specific warm-up before each of the tests. All positions were controlled and recorded at the first 1RM test and reproduced during the RT sessions. Maximal leg press strength was defined as the maximal load lifted in a controlled fashion, starting from a knee angle of 90 degrees. To find a reproducible 90-degree knee angle for each participant, centimetre markings on the side panels of the leg press machine were used to record where to find 90 degrees for each separate leg and participant. Attempts where participants did not reach 90 degrees during the eccentric phase, were not approved. Maximal knee extension strength testing followed the same specific warm-up as the maximal leg press test and was defined as the maximal load lifted in a controlled fashion, reaching full extension of the knee joint. Attempts with exaggerated hip movement or beneath full extension were not approved. Two minutes of rest were given during the specific warm-up, and three minutes of rest were given between 1RM attempts.

@@

Resistance training consisted of three sets of unilateral leg presses and three sets of unilateral knee extensions, with an exercise intensity of 10 repetitions maximum (10RM). As a general warm-up, the participants cycled on an indoor exercise bicycle for 5-10 minutes. In addition, before the respective exercises, two 10-repetition warm-up sets were completed at \~50% and \~70% of 10RM. To ensure adequate exercise stimulation throughout the intervention, the exercise load was increased the following set if the participants lifted more than 12 repetitions, as a progressive loading strategy. If the participants lifted fewer than 8 repetitions per set, the load was reduced in the following set. The resting time between working sets was two minutes. For safety and standardisation purposes, all sessions were monitored by trained personnel. Lastly, training volume (load and repetitions) was logged for every session.

@@

To measure blood glucose levels with and without glucose ingestion/training, capillary blood was collected from finger draws on days with biopsy sampling. One capillary blood sample was collected on day 1 (Figure 1A) to serve as a baseline. On days 11 and 12 (Figure 1A), capillary blood samples were collected seven times: I) Immediately before protein ingestion (0700hrs) II) 45 minutes after protein ingestion (0745hrs) III) 1.5 hours after protein ingestion (0830hrs, i.e., immediately before glucose/placebo intake), IV) 2 hours after protein ingestion (0900hrs, i.e., immediately before training), IV) in the middle of RT (\~0915hrs), V) immediately after training (\~0930hrs), and VI) 2 hours after completion of training (\~1130hrs). Capillary blood samples were analysed with in-house equipment (BIOSEN C-Line, EKF diagnostic GmbH, Barleben). Venous blood samples were collected from the antecubital vein, coinciding with the capillary samples except 45 minutes after protein ingestion and in the middle of the RT session, to analyse endocrine variables.
